Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-30 manuscript/main.tex
minor revisions

Tropical ecosystem services are severely impacted by deforestation and
forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
only does tropical forest Land-Use and Land-Cover Change (LULCC)
constitute 10 to 15\% of the total global carbon emissions
\citep{vanderwerf2009}, changes in forest fragmentation affect the
forest structure and function
\citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
decreases the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.

@@

Historical estimates \textbf{of Land-Use and Land-Cover (LULC)}, in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post World War II German \citep{nita2018}. Survey data across
the African continent is less common, inaccessible or both. Some studies
do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used aerial
survey images to map vegetation changes in South African savannas,
whilst \citet{frankl2015} and \citet{nyssen2016} mapped the Ethiopian
highlands of 1930's.

@@

Here, we use a combination of historical aerial photography and
contemporary remote sensing data to map long-term changes in the extent
and structure of the tropical forest surrounding Yangambi (DR Congo), in
the central Congo Basin, effectively linking the start of the
antrophocene \citep{lewis2015} with current assessments. Yangambi was,
and remains, a focal center of forest and agricultural research and
development in the central Congo Basin. Past research in the region
allows for thorough assessment of land-use and land-cover change from a
multi-disciplinary point of view, confronting us with complex
deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of
historical imagery and develop a Convolutional Neural Network based
forest cover mapping approach, based upon a semi-supervised generated
dataset extensively leveraging data augmentation. Our methodology aims
to provides a historical insight into important LULCC spatial patterns
in Yangambi, such as fragmentation and edge complexity. We further
contextualize the influence of changes in the forest's life history on
past and current research into Above Ground Carbon (AGC) storage
\citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in the
central Congo Basin. Our fast scalable mapping approach for historical
aerial survey data, using limited supervised input, would further
support long-term land-use and land-cover change analysis across the
central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa, DR Congo (then Léopoldville). Black-and-White infrared images
(0.4 - 0.9 \(\mu\)m) were gathered along flight paths running mostly
from west to east, between \textbf{09:00 and 11:00} local time. Along a
flight path continuous images were taken using a Wild Heerburgg RC5a
(currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f
5.6, with a 90\degree view angle) resulting in square photo negatives of
180 by 180 mm. Flights were flown at an average absolute altitude of
\textasciitilde{}5200 m above sea level, covering roughly 18 530
km\textsuperscript{2} at an approximate scale of 1/40 000. The use of
the integrated autograph system ensured timely acquisition of pictures
with a precise overlap (\textasciitilde{}1/3) between images. This large
overlap between images together with flight parameters would allow
post-processing, using stereographs, to create accurate topographic
maps. Original data from this campaign are stored in the
\href{https://www.africamuseum.be/en}{Royal Museum for Central Africa}
in Tervuren, Belgium (Figure 1).

@@

We calculated the orthomosaic using a low resolution point cloud and
digital elevation map (DEM). Additional ground control points were
provided to assist in the referencing of image and constrain the
optimization routine used in the SfM algorithm. Ground control points
consisted of permanent structures which could be verified in both old
and new aerial imagery (i.e.~ESRI World Imagery) \textbf{and consisted
of corner points of build structures (e.g.~a building, bridge or
swimming pool etc.).} Although clouds were removed during the SfM
routine we did not mask all clouds in the final orthomosaic to maximize
forest coverage. The final scene was cropped, to provide consistent wall
to wall coverage of the reconstructed scene. The orthomosaic was
exported as a geotiff for further georeferencing in QGIS
\citep{qgisdevelopmentteam2018} using the georeferencer plugin (version
3.1.9) and additional ESRI World Imagery high resolution reference data.
We used 3\textsuperscript{th} degree polynomial and 16 ground control
points to correct the final image. Ground control points, raw image data
and final processed image are provided in addition to measures of
uncertainty such as mean and median error across all ground control
points. All subsequent analysis are executed on the final geo-referenced
orthomosaic or subsets of it.

@@

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands
which \textbf{showed visible canopy cover loss}, fields and buildings.
We used the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
architecture implemented in Keras \citep{chollet2015} with an
efficientnetb3 pre-processing backbone \citep{yakubovskiy2019} running
on TensorFlow \citep{martinabadi2015} to train a binary classifier
(i.e.~forest or non-forested). Training data were collected from the
orthomosaic by randomly selecting 513 pixel square tiles from
\textbf{locations within} homogeneous forested or non-forested polygons
in the historical orthomosaic \textbf{(Figure 5). Separate polygons were
selected for training, testing and validation purposes. Validation
polygons were sampled 300 times, while both testing and validation
polygons were sampled at 100 random locations. Tiles extracted from
locations close to the polygon border at times contained mixed cover
types. Tiles with mixed cover types were removed from the list of source
tiles (Table 1).} Homogeneous \textbf{source} tiles were combined in
synthetic landscapes using a random gaussian field based binary mask
(Figure 3). We generated 5000 synthetic landscapes for training, while
500 landscapes were generated for both the validation and the testing
datasets for a total of 6000 synthetic landscapes. In order to limit
stitch line misclassifications, along the seams of mosaicked images, we
created synthetic landscapes with different forest tiles to mimick
forest texture transitions. We applied this technique to 10\% of the
generated synthetic landscapes \textbf{(across training, validation and
testing data)}.

@@

The CNN model was trained for 100 epochs with a batch size of 30 using
Adam optimization \citep{kingma2017}, maximizing the
Intersect-over-Union (IoU) using Sørensen--Dice \citep{dice1945} and
categorical cross-entropy loss functions. Data augmentation included
random cropping to 320 pixel squares, random orientation, scaling,
perspective, contrast and brightness shifts and image blurring. The
optimized model was used to classify the complete orthomosaic using a
moving window approach with a step size of 110 pixels and a majority
vote \textbf{(\textgreater{} 50\% agreement)} across overlapping areas
to limit segmentation edge effects. \textbf{In addition, we provide raw
pixel level classification agreement data for quality control purposes
(see Data availability below).} We refer to Figure 4 for a synoptic
overview of the full deep learning workflow.

@@

\textbf{We report the CNN accuracy based upon the IoU of our
out-of-sample validation dataset of synthetic landscapes. In addition,
we report confusion matrices for all pixels across the homogeneous
validation polygons, as well as the training and testing polygons (see
Figure 5). Furthermore, we used the first acquisition of a recent
pan-chromatic Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado,
U.S.A., order 737537, 2011-11-11 08:55 GMT or 09:55 local time) to
classify and assess the robustness of the CNN algorithm on contemporary
remote sensing imagery with similar spectral and spatial
characteristics. We used the Global Forest Change version 1.6 (GFC, tile
10N-020E) \citep{hansen2013} map data, capturing forest loss up to 2011,
to quantify accuracy on downsampled CNN Geo-Eye classification results.
Once more, we report the confusion matrix of between the GFC and CNN
derived forest cover maps, masking clouds and cloud shadows. To
summarize confusion matrices we report accuracy as:}

@@

\textbf{in which TP, TN, FP, FN are True Positive, True Negative, False
Positive and False Negative, respectivelly.}

@@

To map long term LULCC in the Yangambi region we used the contemporary
Global Forest Change version 1.6 (GFC, tile 10N-020E) map data
\citep{hansen2013}. Using the \textbf{GFC} data we calculated the latest
state of the forest with respect to the conditions at the start of 1958,
60 years earlier. In our analysis we only included \textbf{GFC} pixels
which recorded no \textbf{forest} loss throughout the whole 2000 - 2018
period. As the resolution of the historical forest classification
exceeds that of the GFC map we downsampled our historical forest cover
data to 30 m GFC resolution \textbf{using a nearest neighbour approach}.
We masked out all water bodies using the Global Forest Change survey
data mask layer, and limited the analysis to the right bank of the Congo
river. We provide summary statistics of historical and contemporary
forest loss and regrowth. \textbf{We map persistent forest loss after
1958, regrowth after loss in 1958, recent forest loss and long term
(stable) forest cover.}

@@

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape fragmentation
metrics \citep{dale1999}. We report the ratio of edge to area and the
fractal index to quantify landscape complexity of forest disturbances.
Statistics were calculated for all forest disturbance patches larger
than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
patch size distribution using the R package landscapemetrics
\citep{hesselbarth2019}. We provide mean and standard deviation on edge,
area, their ratio and fractal dimension for both the historical and
contemporary \citet{hansen2013} forest cover maps.

We estimated above ground carbon (AGC) losses and gains over time using
plot based averages of recent inventory data at permanent sampling plots
in the area (Figure 2). We refer to \citet{Kearsley2013} for the survey
method and allometric relations used to scale the survey data. Unlike
standard square 1 ha plots, edge plots were set back 200 m from forest
edges and were 50x200 m, with the 50 m side of the plot along the forest
edge and continuing 200 m into the forest (Appendix Table 1). We further
confirmed that forest edge plots did not show a significantly different
AGC compared to those of non-edged / mixed forest plots (Mann Whitney U
test, p \textless{} 0.05). Thus it was not necessary to explicitly
quantify changes in AGC caused by edge effects. Moreover, we used the
mean values of the mixed forest as representative for potential AGC
losses. Despite the challenges inherent in quantifying AGC for forest
edges we mapped the total extent of the edges in the contemporary
landscape. To align our landscape analysis with exploratory analysis of
the survey data we used a buffer of 200 m to estimate the extent of
forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on
age and the crop type. For example, the AGC values varied from 168.67 to
86.55 Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis} (rubber
tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
(\citet{bustillo2018}, personal communications). These higher values are
in line with the mixed AGC estimates in the area, while the palm
plantations resemble old-regrowth values (81.8 Mg C
ha\textsuperscript{-1}, see \textbf{Appendix Table 1}). We therefore use
both the estimates of old-regrowth and mixed forest to estimate AGC for
regrowth. We did not have sufficient data to account for individual
changes in AGC across plantations.

@@

Our analysis provides a first spatially explicity historical composite
of aerial survey images \textbf{in support of mapping land-use and
land-cover} within the Congo Basin. The use of high resolution
historical images combined with SfM image processing techniques allowed
us to mosaic old imagery across a large extent. The final orthomosaic
composition of the Yangambi region \textbf{provided} an image scene
covering approximately 733 million pixels across
\textbf{\textasciitilde{}82800 ha} (\textasciitilde{}23x36 km, Figure
2). The overall accuracy of the SfM orthomosaic composition was 0.88
m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The
resulting georeferenced scene had a spatial accuracy of approximately
\textbf{23 m}. Further georeferencing outside the SfM workflow reduced
the mean error at the ground control points to 5.3 \(\pm\) 4.9 px
(\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px (2.6
m). The orthomosaic served as input for all subsequent LULCC analysis
with all derived maps provided with the manuscript repository (see data
\& code availability statements below).

@@

\textbf{The CNN deep learning classifier reached an
Intersection-over-Union accuracy of 97\% on the detection of disturbed
forest in the out-of-sample (validation) synthetic landscape data. Using
all pixels within the validatoin polygons (Figure 5) showed a similar
accuracy value of \textasciitilde{}98\%. Using all polygons across the
scene, including those used in the generation of testing and training
synthetic landscapes, increased the accuracy to \textasciitilde{}99\%
(Table 2). A comparison with recent pan-chromatic Geo-Eye data shows
good agreement, with an accuracy of \textasciitilde{}87\% across all
pixels, between the landsat based GFC data and downscaled CNN results
(Table 2 and Figure 7).}

@@

\caption{\label{tab:unnamed-chunk-7}Confusion matrix showing \% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas.}
\centering
\begin{tabular}[t]{lrrrrrr}
\toprule
\multicolumn{1}{c}{CNN} & \multicolumn{2}{c}{Geo-Eye (full scene)} & \multicolumn{2}{c}{Aerial Survey (all polygons)} & \multicolumn{2}{c}{Aerial Survey (validation polygons)} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
  & non-forest & forest & non-forest & forest & non-forest & forest\\
\midrule
non-forest & 10.17 & 8.55 & 97.4 & 0.19 & 98.59 & 0.0\\
forest & 3.75 & 77.52 & 2.6 & 99.81 & 1.41 & 100.0\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Accuracy}}\\
\hspace{1em} &  & 87.70 &  & 98.61 &  & 99.3\\
\bottomrule
\end{tabular}
\end{table}

@@

Scaling our classifier to the whole historical orthomosaic we detected
\textbf{\textasciitilde{}16200 ha} (or \textasciitilde{}20\% of the
scene) of disturbed forests. A large fraction of the disturbed area was
restored in the period between the two \textbf{acquisition periods}. In
total, \textbf{9918 ha}, or little over half of the affected forest was
restored (Figure 5C-D, dark blue). Recent deforested areas, as
registered through satellite remote sensing, approximate \textbf{8776 ha
(Table 3, Figure 6 - light green)}.

Recent deforestation follows a distinctly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (\textbf{figure 6C}, dark blue
and green colours respectively). These differences are reflected in the
analysis of landscape metrics of forest loss. Between the historical and
contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to
\textasciitilde{}1.86 \(\pm\) 0.75 ha from \textasciitilde{}5.25 \(\pm\)
5.02 ha historically. Perimeter lengths were longer historically, at
1451 \(\pm\) 943 m, compared to contemporary landscapes
\textasciitilde{}921 \(\pm\) 362 m (Table 3). This shift in perimeter
area ratio led to a similar change in the fractal index, slightly
increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04 over
time. Values closer to a fractal index of 2 suggest a more complex
(fragmented) landscape.

A comparison of forest edge plots with mixed forest plots showed no
signficant difference in AGC, or other reported values such as species
richness, basal area or stem density (Mann Whitney U test, p \textless{}
0.05). Edge influence did not extend beyond 200 m from a forest edge,
but still represented an area of \textbf{13151 ha}.

@@

\caption{\label{tab:unnamed-chunk-10}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
\centering
\begin{tabular}[t]{lrl}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{1}{c}{AGC} \\
\cmidrule(l{3pt}r{3pt}){3-3}
  & ha & Gg C $ha\textsuperscript{-1}$\\
\midrule
Forest & 68455 & \\
-  of which edges & 13151 & \\
Regrowth > 1958 loss & 9918 & 811 - 1592\\
Loss > 2000 & 8776 & 1408\\
Loss > 1958 (permanent) & 6282 & 1008\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery. Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-13}
\end{figure}

>>>>

Tropical ecosystem services are severely impacted by deforestation and
forest degradation \citep{hansen2013, houghton2012, tyukavina2015}. Not
only does tropical forest Land-Use and Land-Cover Change (LULCC)
constitute 10 to 15\% of the total global carbon emissions
\citep{vanderwerf2009}, changes in forest fragmentation affect the
forest structure and function
\citep{fauset2017, brinck2017, didham1999}. Strong fragmentation effects
\textbf{decrease} the number of large trees along forest edges
\citep{laurance2000, magnago2015}, while species composition and
biodiversity are equally negatively affected
\citep{poorter2006, barlow2016, vandeperre2018}. Estimates show that
31\% of carbon emissions are caused by edge effects alone
\citep{brinck2017}.

@@

Historical estimates of Land-Use and Land-Cover (LULC), in the
pre-satellite era (\textless{} 1972), exist but generally rely on
non-spatially explicit data (i.e socio-economic data)
\citep{houghton2012, ramankutty1999, defries2002, sader1988}. Efforts
have been made to use other geo-spatial data sources such as historical
maps \citep{willcock2016}, declassified CORONA satellite surveillance
data across the US and central Brazil \citep{song2015} as well as aerial
surveys in post World War II Germany \citep{nita2018}. Survey data
across the African continent is less common, inaccessible or both. Some
studies do exist, as \citet{buitenwerf2012} and \citet{hudak1998} used
aerial survey images to map vegetation changes in South African
savannas, whilst \citet{frankl2015} and \citet{nyssen2016} mapped the
Ethiopian highlands of \textbf{1930s}.

@@

Here, we use a combination of historical aerial photography
\textbf{(1958)} and contemporary remote sensing data
\textbf{(2000-2018)} to map long-term changes in the extent and
structure of the tropical forest surrounding Yangambi (DR Congo), in the
central Congo Basin, effectively linking the start of the
\textbf{anthrophocene} \citep{lewis2015} with current assessments.
Yangambi was, and remains, a focal center of forest and agricultural
research and development in the central Congo Basin. Past research in
the region allows for thorough assessment of land-use and land-cover
change from a multi-disciplinary point of view, confronting us with
complex deforestation and land-use patterns.

We leverage Structure-from-Motion to generate a large orthomosaic of
historical imagery and develop a Convolutional Neural Network based
forest cover mapping approach, based upon a semi-supervised generated
dataset extensively leveraging data augmentation. Our methodology aims
to \textbf{provide} a historical insight into important LULCC spatial
patterns in Yangambi, such as fragmentation and edge complexity. We
further contextualize the influence of changes in the forest's life
history on past and current research into Above Ground Carbon (AGC)
storage \citep{Kearsley2013} and biodiversity \citep{vandeperre2018} in
the central Congo Basin. Our fast scalable mapping approach for
historical aerial survey data, using limited supervised input, would
further support long-term land-use and land-cover change analysis across
the central Congo Basin.

@@

Data for the central Congo Basin region, surrounding Kisangani, were
collected in several flights during the dry season of 1958 and 1959
(from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959
respectively, see Appendix Figure 1) to generate topographic maps of the
area, supervised by the ``Institut Géographique du Congo Belge'' in
Kinshasa, DR Congo (then Léopoldville). Black-and-White
\textbf{panchromatic} images (0.4 - 0.9 \(\mu\)m) were gathered along
flight paths running mostly from west to east, between 09:00 and 11:00
local time. Along a flight path continuous images were taken using a
Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens
assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in
square photo negatives of 180 by 180 mm. Flights were flown at an
average absolute altitude of \textasciitilde{}5200 m above sea level,
covering roughly 18 530 km\textsuperscript{2} at an approximate scale of
1/40 000. The use of the integrated autograph system ensured timely
acquisition of pictures with a precise overlap (\textasciitilde{}1/3)
between images. This large overlap between images together with flight
parameters would allow post-processing, using stereographs, to create
accurate topographic maps. Original data from this campaign are stored
in the \href{https://www.africamuseum.be/en}{Royal Museum for Central
Africa} in Tervuren, Belgium (Figure 1).

@@

We calculated the orthomosaic using a low resolution point cloud and
digital elevation map (DEM). Additional ground control points were
provided to assist in the referencing of image and constrain the
optimization routine used in the SfM algorithm. Ground control points
consisted of permanent structures which could be verified in both old
and new aerial imagery (i.e.~ESRI World Imagery) and consisted of corner
points of build structures (e.g.~a building, bridge or swimming pool
etc.). \textbf{Although most clouds were removed during the SfM routine
some were retained to provide sufficient SfM tie points to maximize
continuous forest coverage in the final orthomosaic.} The final scene
was cropped, to provide consistent wall to wall coverage of the
reconstructed scene. The orthomosaic was exported as a geotiff for
further georeferencing in QGIS \citep{qgisdevelopmentteam2018} using the
georeferencer plugin (version 3.1.9) and additional ESRI World Imagery
high resolution reference data. We used \textbf{3\textsuperscript{rd}}
degree polynomial and 16 ground control points to correct the final
image. Ground control points, raw image data and final processed image
are provided in addition to measures of uncertainty such as mean and
median error across all ground control points. All subsequent analysis
are executed on the final geo-referenced orthomosaic or subsets of it.

@@

We automatically delineated all natural forest in the historical data,
thus excluding tree plantations, thinned or deteriorated forest stands
which showed visible canopy cover loss, fields and buildings. We used
the Unet Convolutional Neural Net (CNN, \citet{ronneberger2015})
architecture implemented in Keras \citep{chollet2015} with an
efficientnetb3 pre-processing backbone \citep{yakubovskiy2019} running
on TensorFlow \citep{martinabadi2015} to train a binary classifier
(i.e.~forest or non-forested). Training data were collected from the
orthomosaic by randomly selecting 513 pixel square tiles from locations
within homogeneous forested or non-forested polygons in the historical
orthomosaic (Figure 5). Separate polygons were selected for training,
testing and validation purposes. Validation polygons were sampled 300
times, while both testing and validation polygons were sampled at 100
random locations. Tiles extracted from locations close to the polygon
border at times contained mixed cover types. Tiles with mixed cover
types were removed from the list of source tiles (Table 1). Homogeneous
source tiles were combined in synthetic landscapes using a random
gaussian field based binary mask (Figure 3). We generated 5000 synthetic
landscapes (balancing forest and non-forest classes) for training, while
500 landscapes were generated for both the validation and the testing
datasets for a total of 6000 synthetic landscapes. In order to limit
stitch line misclassifications, along the seams of mosaicked images, we
created synthetic landscapes with different forest tiles to mimick
forest texture transitions. We applied this technique to 10\% of the
generated synthetic landscapes (across training, validation and testing
data).

@@

The CNN model was trained for 100 epochs with a batch size of 30 using
Adam optimization \citep{kingma2017}, maximizing the
Intersect-over-Union (IoU) using Sørensen--Dice \citep{dice1945} and
categorical cross-entropy loss functions. Data augmentation included
random cropping to 320 pixel squares, random orientation, scaling,
perspective, contrast and brightness shifts and image blurring. The
optimized model was used to classify the complete orthomosaic using a
moving window approach with a step size of 110 pixels and a majority
vote (\textgreater{} 50\% agreement) across overlapping areas to limit
segmentation edge effects. In addition, we provide raw pixel level
classification agreement data for quality control purposes (see Data
availability below). We refer to Figure 4 for a synoptic overview of the
full deep learning workflow.

@@

We report the CNN accuracy based upon the IoU of our out-of-sample
validation dataset of synthetic landscapes. In addition, we report
confusion matrices for all pixels across the homogeneous validation
polygons, as well as the training and testing polygons (see Figure 5).
Furthermore, we used the first acquisition of a recent pan-chromatic
Geo-Eye 1 stereo pair (Geo-Eye, Thornton, Colorado, U.S.A., order
737537, 2011-11-11 08:55 GMT or 09:55 local time) to classify and assess
the robustness of the CNN algorithm on contemporary remote sensing
imagery with similar spectral and spatial characteristics. We used the
Global Forest Change version 1.6 (GFC, tile 10N-020E) \citep{hansen2013}
map data, capturing \textbf{deforestation} up to 2011, to quantify
accuracy on downsampled CNN Geo-Eye classification results. Once more,
we report the confusion matrix of between the GFC and CNN derived forest
cover maps, masking clouds and cloud shadows. To summarize confusion
matrices we report accuracy as:

@@

in which TP, TN, FP, FN are True Positive, True Negative, False Positive
and False Negative, respectivelly.

@@

To map long term LULCC in the Yangambi region we used the contemporary
Global Forest Change version 1.6 (GFC, tile 10N-020E) map data
\citep{hansen2013}. Using the GFC data we calculated the latest state of
the forest with respect to the conditions at the start of 1958, 60 years
earlier. In our analysis we only included GFC pixels which recorded no
forest loss throughout the whole 2000 - 2018 period, \textbf{as such
locations which would see reforestation or deforestation between 2000
and 2018 would be marked as non-forest (i.e.~disturbed)}. As the
resolution of the historical forest classification exceeds that of the
GFC map we downsampled our historical forest cover data to 30 m GFC
resolution using a nearest neighbour approach. We masked out all water
bodies using the Global Forest Change survey data mask layer, and
limited the analysis to the right bank of the Congo river. We provide
summary statistics of historical and contemporary \textbf{deforestation
and reforestation}. We map permanent forest loss after 1958,
\textbf{reforestation} after 1958, \textbf{recent deforestation} and
long term (stable) forest cover. \textbf{All references to changes over
time in the context of our analysis explicitly compare the historical
and contemporary periods from hereon forth.}

@@

To quantify changes in the structure of forest cover and its
disturbances we used spatial pattern analysis landscape (fragmentation)
metrics \citep{dale1999}. Landscape metrics provide a mathematical
framework for the analysis of discrete land‐cover classes and allows us
to capture their composition and configuration. These metrics are
therefore commonly used to compare how landscapes change over time
\citep{hesselbarth2019}. In particular, fractals provide a way to
quantify complex natural landscapes, including their self-similarity,
across scales \citep{Li2000, Mandelbrot1977} We report the ratio of edge
to area and the fractal dimension to quantify landscape complexity of
forest disturbances. A fractal dimension closer to 2 suggest a more
complex (fragmented) landscape.

Statistics were calculated for all forest disturbance patches larger
than 1 ha and smaller than the 95\textsuperscript{th} percentile of the
patch size distribution using the R package landscapemetrics
\citep{hesselbarth2019}. We provide mean and standard deviation on edge,
area, their ratio and fractal dimension for both the historical and
contemporary \citet{hansen2013} forest cover maps.

We estimated above ground carbon (AGC) losses and gains, \textbf{due to
deforestation and reforestation}, using plot based averages of recent
inventory data at permanent sampling plots in the area (Figure 2). We
refer to \citet{Kearsley2013} for the survey method and allometric
relations used to scale the survey data. Unlike standard square 1 ha
plots, edge plots were set back 200 m from forest edges and were 50x200
m, with the 50 m side of the plot along the forest edge and continuing
200 m into the forest (Appendix Table 1). We further confirmed that
forest edge plots did not show a significantly different AGC compared to
those of non-edged / mixed forest plots (Mann Whitney U test, p
\textless{} 0.05). Thus it was not necessary to explicitly quantify
changes in AGC caused by edge effects. Moreover, we used the mean values
of the mixed forest as representative for potential AGC losses. Despite
the challenges inherent in quantifying AGC for forest edges we mapped
the total extent of the edges in the contemporary landscape. To align
our landscape analysis with exploratory analysis of the survey data we
used a buffer of 200 m to estimate the extent of forest edges and
patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on
age and the crop type. For example, the AGC values varied from 168.67 to
86.55 Mg C ha\textsuperscript{-1}, for \emph{Hevea brasiliensis} (rubber
tree) and \emph{Elaeis guineensis} (oil palm) plots respectively
(\citet{bustillo2018}, personal communications). These higher values are
in line with the mixed AGC estimates in the area, while the palm
plantations resemble old-regrowth values (81.8 Mg C
ha\textsuperscript{-1}, see Appendix Table 1). We therefore use both the
estimates of old-regrowth and mixed forest to estimate AGC for regrowth.
We did not have sufficient data to account for individual changes in AGC
across plantations.

@@

Our analysis provides a first spatially explicity historical composite
of aerial survey images in support of mapping land-use and land-cover
within the Congo Basin. The use of high resolution historical images
combined with SfM image processing techniques allowed us to mosaic old
imagery across a large extent. The final orthomosaic composition of the
Yangambi region provided an image scene covering approximately 733
million pixels across \textasciitilde{}82800 ha (\textasciitilde{}23x36
km, Figure 2). The overall accuracy of the SfM orthomosaic composition
was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8
m/pixel. The resulting georeferenced scene had a spatial accuracy of
approximately 23 m. Further georeferencing outside the SfM workflow
reduced the mean error at the ground control points to 5.3 \(\pm\) 4.9
px (\textasciitilde{}4.7 \(\pm\) 4.3 m), with a median error of 2.9 px
(2.6 m). The orthomosaic served as input for all subsequent LULCC
analysis with all derived maps provided with the manuscript repository
(see data \& code availability statements below).

@@

The CNN deep learning classifier reached an Intersection-over-Union
accuracy of 97\% on the detection of disturbed forest in the
out-of-sample (validation) synthetic landscape data. Using all pixels
within the \textbf{validation} polygons (Figure 5) showed a similar
accuracy value of \textasciitilde{}98\%. Using all polygons across the
scene, including those used in the generation of testing and training
synthetic landscapes, increased the accuracy to \textasciitilde{}99\%
(Table 2). A comparison with recent pan-chromatic Geo-Eye data shows
good agreement, with an accuracy of \textasciitilde{}87\% across all
pixels, between the landsat based GFC data and downscaled CNN results
(Table 2 and Figure 7).

@@

\caption{\label{tab:unnamed-chunk-7}Confusion matrix showing \% agreement between forest / non-forest classes using a Convolutional Neural Network (CNN) across previously selected homogenous areas. In addition, overall accuracy is reported for each confusion matrix.}
\centering
\begin{tabular}[t]{lrrrrrr}
\toprule
\multicolumn{1}{c}{CNN} & \multicolumn{2}{c}{Geo-Eye (full scene)} & \multicolumn{2}{c}{Aerial Survey (all polygons)} & \multicolumn{2}{c}{Aerial Survey (validation polygons)} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
  & non-forest & forest & non-forest & forest & non-forest & forest\\
\midrule
non-forest & 10.17 & 8.55 & 97.4 & 0.19 & 98.59 & 0.0\\
forest & 3.75 & 77.52 & 2.6 & 99.81 & 1.41 & 100.0\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Accuracy}}\\
\hspace{1em} &  & 87.70 &  & 98.61 &  & 99.3\\
\bottomrule
\end{tabular}
\end{table}

@@

Scaling our classifier to the whole historical orthomosaic we detected
\textasciitilde{}16200 ha (or \textasciitilde{}20\% of the scene) of
disturbed forests. A large fraction of the disturbed area was restored
in the period between the two acquisition periods. In total, 9918 ha, or
little over half of the affected forest was restored (Figure 5C-D, dark
blue). Recent deforested areas, as registered through satellite remote
sensing \textbf{(\textgreater{} 2000)}, approximate 8776 ha (Table 3,
Figure 6 - light green).

Recent deforestation follows a distinctly different pattern compared to
historical patterns. Historical deforestation showed a classical
fishbone pattern for forest clearing with very sharp edges, while
current patterns are patchy and ad-hoc (figure 6C, dark blue and green
colours respectively). These differences are reflected in the analysis
of landscape metrics of \textbf{deforestation}. Between the historical
and contemporary LULCC maps we see an increase in small disturbances, as
indicated by the decreasing area of the mean patch size, down to
\textasciitilde{}1.86 \(\pm\) 0.75 ha from \textasciitilde{}5.25 \(\pm\)
5.02 ha historically. Perimeter lengths were longer historically, at
1451 \(\pm\) 943 m, compared to contemporary landscapes
\textasciitilde{}921 \(\pm\) 362 m (Table \textbf{4}). This shift in
perimeter area ratio led to a similar change in the fractal index,
slightly increasing in value to 1.1 \(\pm\) 0.05 from 1.09 \(\pm\) 0.04
over time. Values closer to a fractal index of 2 suggest a more complex
(fragmented) landscape.

A comparison of forest edge plots with mixed forest plots showed no
signficant difference in AGC, or other reported values such as species
richness, basal area or stem density (Mann Whitney U test, p \textless{}
0.05). Edge influence did not extend beyond 200 m from a forest edge,
but still represented an area of 13151 ha (Table 3).

@@

\caption{\label{tab:unnamed-chunk-10}Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.}
\centering
\begin{tabular}[t]{lrl}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{1}{c}{AGC} \\
\cmidrule(l{3pt}r{3pt}){3-3}
  & ha & Gg C $ha\textsuperscript{-1}$\\
\midrule
Non-forest cover (1958) & 16200 & \\
Forest cover & 68455 & \\
-  of which forest edges & 13151 & \\
reforestation >1958 & 9918 & 811 - 1592\\
deforestation >2000 & 8776 & 1408\\
\addlinespace
deforestation >1958 (permanent) & 6282 & 1008\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{RGB visualizations of the first three principal components of scene wide FOTO texture analysis of historical and current (Geo-Eye) imagery. Current permanent sampling plots of mono-dominant, mixed, fallow and young (edge) forest plots are marked with open triangles, open circles, open squares and crosses, respectively.}\label{fig:unnamed-chunk-14}
\end{figure}
